Accessing the nucleon tensor structure in inclusive deep inelastic scattering

### 2016-07-26 1607 - JMC for g2/Paper/JMC_dis.tex

The first term is also known as the Wandzura-Wilczeck function $g_2^{WW} =
g_1^*(x)$ , and contains all the ``pure twist-2'' chiral even contributions to
the $g_2$ structure coming from quark-quark correlators. The second and third
terms contain all ``pure twist-3'' contributions, i.e., those coming from
quark-gluon-quark correlators. The fourth and fifth terms depend on the
transversity parton distribution function, $h_1$. 
The former is usually neglected for
light quarks since it is proportional to $\mq=O$(1 MeV). In the latter term,
new in our analysis, the transversity distribution is multiplied by a constant
of $O$(100 MeV), and cannot be a priori neglected.

It is important to estimate the size of the variou scontributions to the non Wandzura-Wilczek part of $g_2$. We define the shorthand notation
\begin{align}
g_2^{tw3} & = \frac{1}{2}\,\sum_a e_a^2
    \biggl(
    \widetilde g_T^{a \star}(x) 
    + \int_x^1\frac{dy}{y} \widehat{g}_T^q(y) 
    \biggr) 
&
g_2^{\text{quark}} &= \frac{1}{2}\,\sum_a e_a^2 
 \frac{\mq}{\Lambda} (h_1^q/x)^\star(x),
&
g_2^{\text{jet}} &= \frac{1}{2}\,\sum_a e_a^2 
\frac{\mj-\mq}{\Lambda} \frac{h_1^q(x)}{x}. 
\end{align} 
These are comapred in Figure~\ref{f:g2contrib} to the $g_2-g_2^{WW}$ function
obtained in the very recent JAM15 fit of polarized DIS asymmetries
\cite{Sato:2016tuz}, that includes a large amount of precise data at large $x$
from Jefferson Lab, and simultaneously fits the higher-twist components of in
$g_1$ and $g_2$ to the data. For the ``pure twist-3'' contribution,
$g_2^{tw3}$, {\it i.e.}, the contribution from quark-gluon-quark matrix
elements, we show a model calculation by Braun et al.~\cite{Braun:2011aw}; for
other estimates, see \cite{Other-tw3-calcs}. To estimate the contributions
from quark ($g_2^q$) and jet mass ($g_2^{jet}$) effects, that depend on chiral
odd quark-quark matrix elements, we use the recent Pavia15 fit of the
transversity distribution from Ref.~\cite{Radici:2015mwa}, which is comparable
also to other 
extractions~\cite{Anselmino:2013vqa,Kang:2015msa}. Furthermore, we choose the
values of the mass parameters to be $\mq=5$ MeV and $\mj = 100$ MeV.

@@

It is interesting to consider the moments of the non Wandzura-Wilczek contribution to $g-2$,
\begin{align}
  d_N = N \int_0^1 x^N \big( g_2(x) - g_2^{WW}(x) \big) \ .
\end{align}
For a generic function $f$, let us define it's $N$-th moment as $f[N]=\int_0^1 dx\, x^{N} f(x)$. It is then straightforward to verify that $f^*[N] = N/(N+1) f[N]$ and  
\begin{align}
  d_N & \equiv (N+1) g_2[N] + N g_1[N] \\
  & = \frac12 \sum_q e_q^2 \bigg( N \tilde g_T^q[N] + \hat g_T^q[N]
    + \frac{(N+1) M_q-m_q}{\Lambda} h_1^q[N] \bigg) \ .
\end{align}
\AAcom{The notation used in JAM15 with $x^{N-1}$ for the Nth moment is pretty awkward here. Should we use $x^N$ instead?}

The zero-th moment corresponds to an extension of the Burkhardt-Cottingham sum
rule \cite{Burkhardt:1970ti,Jaffe:1996zw}, that appears to be broken by the
non-perturbative spin-flip contribution from the jet function:  
\begin{align}
  \label{eq:BC}
  \int dx\, g_2(x) = \frac{\mj-\mq}{\Lambda} \int dx\, \frac{1}{x} h_1(x) \ .
\end{align}
This sum rule, in which the pure twist-3 part does not take part, is a
consequence of the Lorentz invariance of the DIS cross section, that entails
$\int_0^1 dx g_1^a(x) = \int_0^1 g_T^q(x)$. It explicitly displays a
contribution from spin-flip processes, that are included in the original
derivation by Burkhardt and Cottingham \cite{Burkhardt:1970ti} but do not show
up in treatments that only consider free field quark propagators for the
struck quark \cite{Jaffe:1996zw}.

Assuming the Burkhardt-Cottingham sum rule to be strictly valid, we obtain a strong constraint on the transversity function,
\begin{align}
   \int dx\, \frac{1}{x} h_1(x) = 0 \ .
\label{eq:ABsumrule}
\end{align}
Even if the BC sun rule is broken by a 
$J = 0$ fixed pole with non-polynomial residue, i.e., if $g_2$ integrates to a
finite but non-zero number \cite{Jaffe:1996zw}), we obtain that $h_1(x)/x$
must be integrable. This entails a bound on the small $x$ behavior of the
transversity, 
\begin{align}
  h_1(x) \propto x^\epsilon \ \ \ \epsilon>0 \ .
\label{eq:ABbound}
\end{align}
This bound will be very useful, \eg, in transversity fits, where the data at
small $x$ is, as yet, very limited, and in general for proper extrapolations
when calculating moments.

The first moment is the first in which a contribution from the pure twist-3 part of $g_2$ appears:
\begin{align}
  d_1 & = \frac12 \sum_q e_q^2 \bigg( 2 \tilde g_T^q[1] + \hat g_T^q[1]
    + \frac{2M_q-m_q}{\Lambda} h_1^q[1] \bigg)
\end{align}
where $h_1^q[1] = \int_0^1 dx $ is the contribution of a quark $q$ to the
tensor charge. In this case, the transevrsity contribution is a background to
the extraction of the poure twist-3 piece. Fortunately, it is a quantity that
can be extracted from the lattice \cite{Green:2012ej,Bali:2014nma,Bhattacharya:2015wna,Abdel-Rehim:2015owa,Bhattacharya:2016zcn} or fitted
\cite{Radici:2015mwa,Anselmino:2015sxa,Kang:2015msa}. Furthermore the new sum rule
\eqref{eq:ABsumrule} and bound \eqref{eq:ABbound} promise to improve future
transversity fits. What is less obvious is how to calculate $M_q-m_q$ from
first principles, since this is related to the spectral function of the quark
propagator. We will, however, discuss later perspectives on how to measure
it. Therefore the pure twist-3 part can, in principle, be properly isolated.

The third moment is interesting because the pure twist-3 part can be related
to quark-gluon-quark correlators, see \cite{Jaffe:1996zw}, and interpreted as
as the average color force experienced by the struck quark as it exits the
nucleon~\cite{Burkardt:2012sd}.  As in the previous case, however, the
transversity contribution needs to be subtracted: 
\begin{align}
  d_2 & = \frac12 \sum_q e_q^2 \bigg( 3 \tilde g_T^q[2] + \hat g_T^q[2]
    + \frac{3M_q-m_q}{\Lambda} h_1^q[2] \bigg)
\end{align}

>>>>

The first term is also known as the Wandzura-Wilczeck function $g_2^{WW} =
-g_1^*(x)$ , and contains all the ``pure twist-2'' chiral even contributions to
the $g_2$ structure coming from quark-quark correlators. The second and third
terms contain all ``pure twist-3'' contributions, i.e., those coming from
quark-gluon-quark correlators. The fourth and fifth terms depend on the
transversity parton distribution function, $h_1$. 
The fourth term is usually neglected for
light quarks since it is proportional to $\mq=O$(1 MeV). In the last term,
new in our analysis, the transversity distribution is multiplied by a constant
of $O$(100 MeV), and cannot be a priori neglected.

It is important to estimate the size of the various contributions to the non Wandzura-Wilczek part of $g_2$. We define the shorthand notation
\begin{align}
g_2^{tw3} & = \frac{1}{2}\,\sum_a e_a^2
    \biggl(
    \widetilde g_T^{a \star}(x) 
    + \int_x^1\frac{dy}{y} \widehat{g}_T^q(y) 
    \biggr) 
&
g_2^{\text{quark}} &= \frac{1}{2}\,\sum_a e_a^2 
 \frac{\mq}{\Lambda} (h_1^q/x)^\star(x),
&
g_2^{\text{jet}} &= \frac{1}{2}\,\sum_a e_a^2 
\frac{\mj-\mq}{\Lambda} \frac{h_1^q(x)}{x}. 
\end{align} 
These are comapred in Figure~\ref{f:g2contrib} to the $g_2-g_2^{WW}$ function
obtained in the very recent JAM15 fit of polarized DIS asymmetries
\cite{Sato:2016tuz}, that includes a large amount of precise data at large $x$
from Jefferson Lab, and simultaneously fits the higher-twist components of in
$g_1$ and $g_2$ to the data. For the ``pure twist-3'' contribution,
$g_2^{tw3}$, {\it i.e.}, the contribution from quark-gluon-quark matrix
elements, we show a model calculation by Braun et al.~\cite{Braun:2011aw}; for
other estimates, see \cite{Other-tw3-calcs}. To estimate the contributions
from quark ($g_2^q$) and jet mass ($g_2^{jet}$) effects, that depend on chiral
odd quark-quark matrix elements, we use the recent Pavia15 fit of the
transversity distribution from Ref.~\cite{Radici:2015mwa}, which is comparable
also to other 
extractions~\cite{Anselmino:2013vqa,Kang:2015msa}. Furthermore, we choose the
values of the mass parameters to be $\mq=5$ MeV and $\mj = 100$ MeV.

@@

It is interesting to consider the moments of the non Wandzura-Wilczek contribution to $g_2$,
\begin{align}
  d_N \equiv (N+1) \int_0^1 x^N \bigg( g_2(x) - g_2^{WW}(x) \bigg) \ .
\end{align}
For a generic function $f$, let us define it's $N$-th moment as $f[N]=\int_0^1 dx\, x^{N} f(x)$. It is then straightforward to verify that $f^*[N] = N/(N+1) \times f[N]$ and  
\begin{align}
  d_N & = (N+1) g_2[N] + N g_1[N] \\
  & = \frac12 \sum_q e_q^2 \bigg( N \tilde g_T^q[N] + \hat g_T^q[N]
    + \frac{(N+1) M_q-m_q}{\Lambda} h_1^q[N] \bigg) \ .
\end{align}

The zero-th moment provides an interesting relationship between the longitudinal and tensor spin structure of the target,
\begin{align}
  \label{eq:BC}
  \int dx\, g_2(x) = \sum_q e_q^2 \frac{\mj-\mq}{\Lambda} \int dx\, \frac{1}{x} h_1^q(x) \ ,
\end{align}
where we used the fact that $\hat g_T^q[0]=0$ identically due to the symmetry properties of the quark-gluon-quark correlators. The sum rule \eqref{eq:BC}, in which the pure twist-3 part does not take part, is a consequence of the Lorentz invariance of QCD interactions, that entails $\int_0^1 dx g_1^a(x) = \int_0^1 g_T^q(x)$. It explicitly displays a contribution from spin-flip processes, that are included in the original derivation by Burkhardt and Cottingham \cite{Burkhardt:1970ti} but do not show up in treatments that only consider free field quark propagators for the struck quark \cite{Jaffe:1996zw}. Assuming the Burkhardt-Cottingham sum rule $\int_0^1 dx\, g_2(x) =0$, that is derived at the cross section level (under suitable assumptions on teh high-energy behavior of te hspin-flip amplitude) rather than from an analysis of collinear correlators, as done here, to be strictly valid, we obtain a strong constraint on the transversity function,
\begin{align}
   \int dx\, \frac{1}{x} h_1^q(x) = 0 \ .
\label{eq:ABsumrule}
\end{align}
Interestingly, one can show that this constraint, if valid at any given scale $Q_0$ is conserved through QCD evolution. Even if the BC sum rule is broken either by a $J = 0$ fixed pole with non-polynomial residue \cite{Jaffe:1996zw} or explicitly as, \eg, in the quark target model \cite{Metz-Kundu} -- i.e., if $g_2$ integrates to a finite but non-zero number -- we obtain that $h_1(x)/x$ must be integrable. This entails a bound on the small $x$ behavior of the transversity,
\begin{align}
  h_1^q(x) \propto x^\epsilon \ \ \ \epsilon>0 \ .
\label{eq:ABbound}
\end{align}
This bound will be very useful, \eg, in transversity fits, where the data at
small $x$ is, as yet, very limited, and in general for proper extrapolations
when calculating moments.

The first moment is the first in which a contribution from the pure twist-3 part of $g_2$ appears:
\begin{align}
  d_1 & = \frac12 \sum_q e_q^2 \bigg( 2 \tilde g_T^q[1] + \hat g_T^q[1]
    + \frac{2M_q-m_q}{\Lambda} h_1^q[1] \bigg)
\end{align}
where $h_1^q[1] = \int_0^1 dx h_1^q(x)$ is the contribution of a quark $q$ to the tensor charge. 
The third moment is also interesting because the pure twist-3 part can be related to quark-gluon-quark correlators, see  \cite{Jaffe:1996zw}, and interpreted as as the average color force experienced by the struck quark as it exits the nucleon \cite{Burkardt:2012sd}:
\begin{align}
  d_2 & = \frac12 \sum_q e_q^2 \bigg( 3 \tilde g_T^q[2] + \hat g_T^q[2]
    + \frac{3M_q-m_q}{\Lambda} h_1^q[2] \bigg)
\end{align}
In both of these cases, the transversity contribution is a background to the extraction of the pure twist-3 piece. Fortunately, it is a quantity that can be extracted from the lattice
\cite{Green:2012ej,Bali:2014nma,Bhattacharya:2015wna,Abdel-Rehim:2015owa,Bhattacharya:2016zcn} or fitted
\cite{Radici:2015mwa,Anselmino:2015sxa,Kang:2015msa}.
Furthermore, the new sum rule \eqref{eq:ABsumrule} and the bound \eqref{eq:ABbound} promise to improve future transversity fits. What is less obvious is how to calculate $M_q-m_q$ from first principles, since this is related to the spectral function of the quark propagator. We will, however, briefly discuss perspectives on how to measure it. Therefore the pure twist-3 part can, in principle, be properly isolated.

\begin{figure}[tbh]
  \centering
  \parbox[b]{8cm}{\centering
  \includegraphics
      [width=7cm,clip=true]
      {epl_emn_h_X.eps}
  }
  \parbox[b]{8cm}{\vskip-0.5cm\centering
  \includegraphics
      [width=7cm,clip=true]
      {epl_emn_h_h_X.eps}
  }
  \caption{\small
	Single hadron {\it (left)} and double hadron {\it (right)}
	production in $e^+ e^-$ collisions at LO with jet and
	fragmentation correlators.
  }
  \label{fig:epl_emn_h}
\end{figure}

A promising avenue to experimentally access jet functions is, however, through inclusive single hadron production, $e^+ e^- \to h X$, and inclusive dihadron 
production from the same hemisphere, $e^+ e^- \to h h X$, see 
Fig.~\ref{fig:epl_emn_h}.
In single-hadron production, the fragmentation functions $D_h$
play the role of PDFs in DIS, and couple to the jet functions
in an analogous way.
In double hadron production the enlarged number of Dirac structures
of the dihadron fragmentation correlator $D_{2h}$ allows one to access
the jet function in novel ways, and in particular to isolate the
contribution from the helicity-flip $J_0$ term.
Studying and classifying all the possibilities offered by single and
double hadron production will open up a rich phenom, which will in turn
be needed to extract pure twist-3 matrix elements from the $g_2$ structure function, and more in general to perform precise jet mass corrections in DIS.
